Cartograms with Dynamic Features

### 2022-04-23 paper/GHRVis.tex
draft: review

\begin{abstract}

\end{abstract}

>>>>

\begin{abstract}
Cartograms are hybrid representations of geographical and abstract data based on a value-by-area mapping technique. Based on the Dorling cartogram, the Demers cartogram uses squares instead of circles to represent regions, enabling the comparison of regions in a more intuitive way with a higher screen space utilization. One of the disadvantages of the Dorling cartogram and its variants is that regions may be displaced from their original positions, reducing the legibility, readability, and geographical accuracy. To address that, we propose a novel layout algorithm that integrates dynamic topological features such as rivers into Demers cartograms. We evaluate the proposed layout algorithm through a user study with an application to an Electronic Health Records (EHR) dataset. The results demonstrate that the proposed layout algorithm is able to preserve some of the original properties of the cartogram, while improving the legibility, readability, and geographical accuracy of the cartogram. 
\end{abstract}

### 2022-04-23 paper/section/data.tex
draft: review

\bobgraph{Clinical Commissioning Groups Shapefile:} Clinical Commissioning Groups (CCGs) are the primary administrative and geographic unit of the National Health Service (NHS) in the UK \cite{nhsNHS}. The number of CCGs changes over time due to the NHS re-organization. The most up-to-date shapefile is available from the Open Geography portalx \cite{opengeographyportalxOpen}. We decided to use the CCG shapefile from 2020 at the time of writing, because there is still no public EHR data published based on the latest CCG re-organization that took place in 2021.

@@

After acquiring the shapefiles, we used QGIS \cite{qgisWelcome} to manually adjust projections, translations and scales, and merge them into one shapefile. Finally, mapshaper \cite{blochMapshaper} is used to convert the merged shapefile into a TopoJSON \cite{TopoJSON} file. TopoJSON eliminates redundant coordinates in the data, this improves the rendering speed of our implementation.

>>>>

\bobgraph{Clinical Commissioning Groups Shapefile:} Clinical Commissioning Groups (CCGs) are the primary administrative and geographic unit of the National Health Service (NHS) in the UK \cite{nhsNHS}. The number of CCGs changes over time due to the NHS re-organization. The most up-to-date shapefile is available from the Open Geography portalx \cite{opengeographyportalxOpen}. We decided to use the CCG shapefile from 2020 at the time of writing, due to the absence of public EHR data published based on the latest CCG re-organization that took place in 2021.

@@

After acquiring the shapefiles, we used QGIS \cite{qgisWelcome} to manually adjust projections, translations and scales, and merge them into one shapefile. Finally, mapshaper \cite{blochMapshaper} is used to convert the merged shapefile into a TopoJSON \cite{TopoJSON} file. TopoJSON eliminates redundant coordinates in the data, improving the rendering speed of our implementation.

### 2022-04-23 paper/section/main.tex
draft: review

\algoref{alg:UpdateNodePosition} and \figref{fig:flowchart} provide an overview of the process. \textbf{Initialization:} We first load and render the CCG geospatial boundaries. For each CCG we compute the centroid and represent it using a small unit size square node with initial size, $ \nodeSize $. We start with small unit size squares because they do not overlap initially. \textbf{Node layout:} The basic algorithm is as follows: we first apply the Fast Node Overlap Removal (FNOR) algorithm that solves the Variable Placement with Separation Constraints (VPSC) problem \cite{dwyer2006fast} in order to remove overlaps. We chose FNOR over other node overlap removal algorithms because FNOR is able to provide spread minimization and node movement minimization while maintaining a good level of global shape preservation \cite{chen2020Node}. We gradually increase the node size by 1 unit at a time to ensure smooth transitions. An increase in $ \nodeSize $ can cause the nodes to overlap. During overlap removal, we compute node trajectories (See \algoref{alg:check river intersection}) and translate nodes to their new position. Nodes that cross a river are translated back to their previous position. If a node oscillates across a river, we process this as a stalemate. This procedure ends when 1) no node overlap is present; 2) no nodes cross a river. We then increase $ \nodeSize $ by one pixel and repeat the algorithm until the max node size $ \nodeSizeMax $ is reached. The gradual size increase process provides stability to the layout and helps minimize geographical error.

@@

We use rivers as topological boundaries and prevent nodes from crossing them. The resolution of rivers can be adjusted by the user, as shown in \figref{fig:river resolution}, the number of edges is reduced from its original 10,170 to 65. We do not require a pixel-level river resolution as Demers cartograms do not need to represent geography at this level od detail. Instead a resolution that matches the node size is used. Enabling river simplification reduces the number of edge intersection tests that need to be performed. When a node's position changes, we test if the node's trajectory intersects any segment of a river. See \algoref{alg:check river intersection}. A bounding box intersection test between the edge defined by node translation and river edges can be performed to reduce the number of edge intersection tests required.

@@

As the FNOR always attempts to produce an optimal node layout where node distribution and translation are minimized, a node's translation path might repeatedly intersect a river due to congestion, creating a stalemate situation, as shown in \figref{fig:stalemate}. If a node is translated between two positions, $ \node $ and $ \nodeFNOR $, for $ \stalemateMax $ iterations (a user-adjustable parameter), we introduce a heuristic solution: constructing a corridor to break up the congestion. A corridor, $ c $, is a rectangle with a width of $ \CorridorWidth $ and a length of $ \CorridorLength $, formed by deriving two edges $ \EdgeParallelA $ and $ \EdgeParallelB $ such that $ \EdgeParallelA \parallel \EdgeParallelA \parallel \nodeLineNtNc $ (See \figref{fig:corridor}C and D). All enclosed nodes are then translated by $ \nodeVectorTC $ to alleviate the congestion (See \figref{fig:corridor}D).

>>>>

\algoref{alg:UpdateNodePosition} and \figref{fig:flowchart} provide an overview of the process. \textbf{Initialization:} We first load and render the CCG geospatial boundaries. For each CCG we compute the centroid and represent it using a small unit size square node, $ \node $, with initial size, $ \nodeSize $. We start with small unit size squares because they do not overlap initially. \textbf{Node layout:} The basic algorithm is as follows: we first apply the Fast Node Overlap Removal (FNOR) algorithm that solves the Variable Placement with Separation Constraints (VPSC) problem \cite{dwyer2006fast} in order to remove overlaps. We chose FNOR over other node overlap removal algorithms because FNOR is able to provide spread minimization and node movement minimization while maintaining a good level of global shape preservation \cite{chen2020Node}. We gradually increase the node size by 1 unit at a time to ensure smooth transitions. An increase in $ \nodeSize $ can cause the nodes to overlap. During overlap removal, we compute node trajectories (See \algoref{alg:check river intersection}) and translate nodes to their new position. Nodes that cross a river are translated back to their previous position. If a node oscillates across a river, we process this as a stalemate. This procedure ends when 1) no node overlap is present; 2) no nodes cross a river. We then increase $ \nodeSize $ by one pixel and repeat the algorithm until the max node size $ \nodeSizeMax $ is reached. The gradual size increase process provides stability to the layout and helps minimize geographical error.

@@

We use rivers as topological boundaries and prevent nodes from crossing them. The resolution of rivers can be adjusted by the user, as shown in \figref{fig:river resolution}, the number of edges is reduced from its original 10,170 to 65. We do not require a pixel-level river resolution as Demers cartograms do not need to represent geography at this level of detail. Instead a resolution that matches the node size is used. Enabling river simplification reduces the number of edge intersection tests that need to be performed. When a node's position changes, we test if the node's trajectory intersects any segment of a river. See \algoref{alg:check river intersection}. A bounding box intersection test between the edge defined by node translation and river edges can be performed to reduce the number of edge intersection tests required.

@@

As the FNOR always attempts to produce an optimal node layout where node distribution and translation are minimized, a node's translation path can repeatedly intersect a river due to congestion, creating a stalemate situation, as shown in \figref{fig:stalemate}. If a node is translated between two positions, $ \node $ and $ \nodeFNOR $, for $ \stalemateMax $ iterations (a user-adjustable parameter), we introduce a heuristic solution: constructing a corridor to alleviate the congestion. A corridor, $ \Corridor $, is a rectangle with a width of $ \CorridorWidth $ and a length of $ \CorridorLength $, formed by deriving two edges $ \EdgeParallelA $ and $ \EdgeParallelB $ such that $ \EdgeParallelA \parallel \EdgeParallelA \parallel \nodeLineNtNc $ (See \figref{fig:corridor}C and D). All nodes enclosed by $ \Corridor $ are then translated by $ \nodeVectorTC $ to alleviate the congestion (See \figref{fig:corridor}E).

### 2022-04-23 paper/section/related.tex
draft: review

Our work extends the algorithm described by \citea{tong2018cartograms} which incorporates a static topological feature into Demers cartograms. We introduce multiple features (rivers) into the layout (as opposed to a single). Making topological features dynamic further improves the legibility and geographical accuracy.

The way we evaluate the cartograms is also different.  \citeauthor{tong2018cartograms} count river crossings to evaluate error. Here our focus is on readability, thus we include a preliminary user study.

>>>>

Our work extends the algorithm described by \citea{tong2018cartograms} which incorporates a static topological feature into Demers cartograms. Our work enhances that of \citeauthor{tong2018cartograms} in three ways. We introduce multiple features (rivers) into the layout (as opposed to a single), making topological features dynamic further improves the legibility and geographical accuracy. We improve the algorithm for resolving stalemates.  The way we evaluate the cartograms is also different. \citeauthor{tong2018cartograms} count river crossings to evaluate error. Here our focus is on readability, thus we include a preliminary user study.
